---
title: "Are Leveraged ETFs Worth It For Young Investor's?"
author: "Trey Davidson"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

```{r, include=FALSE }


update_time_data <- function(data, ticker)
{
  data$Year <- as.numeric(substr(data$Date, 1, 4))
  data$Month <- as.numeric(substr(data$Date, 6, 7))
  data$Day <- as.numeric(substr(data$Date, 9, 10))
  data$DecimalYear <- data$Year + ((data$Month - 1) / 12)
  data$Ticker <- ticker
  
  return(data)
}

update_for_base_year <- function(data1, baseTime) {
 
  base_Open <- data1$Open[data1$DecimalYear == baseTime]
  base_high <- data1$High[data1$DecimalYear == baseTime]
  base_close <- data1$Close[data1$DecimalYear == baseTime]
  base_low <- data1$Low[data1$DecimalYear == baseTime]
  
  
  data1$Open <- (data1$Open / base_Open) * 10000
  data1$High <- (data1$High / base_high) * 10000
  data1$Close <- (data1$Close / base_close) * 10000
  data1$Low <- (data1$Low / base_low) * 10000

  
  return(data1)
}


gen_daily_leveraged_returns <- function(prices, leverage) {
  # Initialize vectors
  leveraged_returns <- numeric(length(prices))
  value <- numeric(length(prices))
  
  # Set the initial value
  value[1] <- prices[1]
  
  # Loop through each day
  for (i in 2:length(prices)) {
    # Calculate the daily leveraged return
    leveraged_returns[i] <- ((prices[i] - prices[i - 1]) / prices[i - 1]) * leverage
    
    
    value[i] <- value[i - 1] * (1 + leveraged_returns[i])
  }
  
  return(value)
}

calculate_daily_returns <- function(prices) {
  returns <- numeric(length(prices))
  returns[1] <- 0  # Corrected spelling here
  for (i in 2:length(prices)) {
    returns[i] <- (prices[i] - prices[i - 1]) / prices[i - 1]
  }
  
  return(returns)
}


# Returns as a decimial
percent_change <- function(new, old)
{
  return((new - old) / old)
}

calculate_CAGR <- function(beg_value, end_value, years) {
  beg_value <- as.numeric(beg_value)
  end_value <- as.numeric(end_value)
  years <- as.numeric(years)
  

  return ((end_value / beg_value)^(1 / years) - 1)
}




```


#### Introduction

I am a young investor who is currently interested in saving for retirement. As a young investor traditional investment advise would say I can afford to take on risk as I have a long time horizon to make up losses through income generation and other methods. If I lose all of my investment at the age of 30, I am essentially equal in position to someone who starts saving money in there 30's. [Considering the fact that 31 is the average age to start saving for retirement I wouldn't be alone](https://news.nationwide.com/americas-retirement-voice-reveals-how-much-it-costs-to-delay-retirement/).


#### The Decesion
Since, I am less concerned about risk I have two choices when it comes to magnifying gains, I can either concentrate a portfolio or take on leverage. The issue with concentration is that many biases can come at play that make stock picking hard unlike leverage, which would allow me to keep the diversity of the S&P 500 or QQQ, while also magnifying returns.



Here are some leveraged ETF's that may be included and there relevant websites for any additional information you may be looking for. 

[SPY](https://www.ssga.com/us/en/intermediary/etfs/spdr-sp-500-etf-trust-spy) - Top 500 Companies US    
[QQQ](https://www.invesco.com/qqq-etf/en/home.html) - Top 100 NASDAQ companies tech weighted heavily    
[UPRO](https://www.proshares.com/our-etfs/leveraged-and-inverse/upro) - 3x Leveraged S&P 500       
[TQQQ](https://www.proshares.com/our-etfs/leveraged-and-inverse/tqqq) - 3x Leveraged QQQ    
[QLD](https://www.proshares.com/our-etfs/leveraged-and-inverse/qld) - 2x Leveraged QQQ    
[SSO](https://www.proshares.com/our-etfs/leveraged-and-inverse/sso) - 3x Leveraged S&P 500    



# Comapring QQQ to QLD

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold'}

qld_data <-  update_time_data(read.csv("QLD.csv"), "QLD")
qqq_data <- update_time_data(read.csv("QQQ.csv"), "QQQ")
qld_data <-  update_for_base_year(qld_data[qld_data$DecimalYear >= 2006.5, ], 2006.5)
qqq_data <- update_for_base_year(qqq_data[qqq_data$DecimalYear >= 2006.5, ], 2006.5)

qld_CAGR <- calculate_CAGR(qld_data$Open[1], qld_data$Open[nrow(qld_data)], 2025 - 2006.5)
qqq_CAGR <- calculate_CAGR(qqq_data$Open[1], qqq_data$Open[nrow(qld_data)], 2025- 2006.5)





ggplot() +
  geom_line(
    data = update_for_base_year(qld_data[qld_data$DecimalYear >= 2006.5, ], 2006.5), 
    aes(x = DecimalYear, y = Close, color = "QLD"), 
    size = 1
  ) +
  geom_line(
    data = update_for_base_year(qqq_data[qqq_data$DecimalYear >= 2006.5, ], 2006.5), 
    aes(x = DecimalYear, y = Close, color = "QQQ"), 
    size = 1
  ) +
  labs(
    title = "QLD vs QQQ",
    x = "Time (Days)",
    y = "Price",
    color = "Stocks"  # This label will appear in the legend
  ) +
  scale_color_manual(
    values = c("QLD" = "blue", "QQQ" = "green"),
    labels = c(
      paste("QLD (CAGR:", round(qld_CAGR * 100, 2), "%)"),
      paste("QQQ (CAGR:", round(qqq_CAGR * 100, 2), "%)")
    )
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )


qld_data$daily_returns <- c(0, diff(log(qld_data$Close)))
qqq_data$daily_returns <- c(0, diff(log(qqq_data$Close)))
```

In our initial backtest, QLD (the 2x leveraged QQQ) significantly outperforms QQQ. However, before drawing any conclusions, it is crucial to assess whether these observed differences could be attributed to random chance. To determine the appropriate statistical test, we need to address two key questions:

   1. Are the monthly returns independent?
   2. Are the monthly returns normally distributed?
   
Regarding question 1, a typical approach would be to use a chi-square test to examine the independence of the returns. However, since QLD is a leveraged derivative of QQQ, we acknowledge that QLD's performance is inherently dependent on QQQ's returns. As such, we will account for this relationship in our analysis.

For question 2, it is important to first visualize the distribution of returns. A histogram can reveal the shape and skewness of the returns, providing insights into their distribution. 

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold'}

ggplot() +
  geom_histogram(data = qld_data, aes(x = daily_returns), 
                 fill = "blue", alpha = 0.5, bins = 50) + 
  geom_histogram(data = qqq_data, aes(x = daily_returns), 
                 fill = "green", alpha = 0.5, bins = 50) + 
  labs(
    title = "Distribution of Monthly Returns: QLD vs QQQ",
    x = "Daily Return",
    y = "Frequency",
    fill = "Stock"
  ) +
  scale_fill_manual(values = c("blue", "green"), labels = c("QLD", "QQQ")) +
  theme_minimal()

```

Based on our observations, the returns are not symmetrical and show a rightward bias, which is favorable for investors, as it indicates a higher frequency of positive returns compared to negative ones. To statistically prove our observation we will use the Shaprio-Wilks test. 


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold'}

test1 <- shapiro.test(qld_data$daily_returns)


test2 <- shapiro.test(qqq_data$daily_returns)


cat("QLD Test - W:", test1$statistic, "p-value:", test1$p.value, "\n")


cat("QQQ Test - W:", test2$statistic, "p-value:", test2$p.value, "\n")


```

The results of the Shapiro-Wilk test confirm our visual analysis, indicating that neither of these distributions is normally distributed. Therefore, we cannot use the t-test, as it assumes normality. Instead, we will use the Wilcoxon test, which does not assume normality or independence of observations

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold'}


test_result <- wilcox.test(qld_data$daily_returns, qqq_data$daily_returns)


cat("Test - W:", test_result$statistic, "p-value:", test_result$p.value, "\n")




```

The Wilcoxon test shows that there is only a 3% probability that the observed differences in returns between QLD and QQQ are due to randomness. This provides sufficient evidence to conclude that QLD had statistically significantly higher returns from 2008 to 2024.

Looking at the returns, we see that an investment of 10K in QQQ grows to 152,537.6, while the same 10K in QLD grows to 547,665.7, which is more than 5 times the return of QQQ, not just 2x. This raises the question: how does a 2x leveraged ETF generate nearly 5x the return of the underlying? 

#### 2x leverage results in 5x the results! 


This happens due to the way percentages stack on top of each other. For example, lets compare two stocks. One that goes up by 50% and then drops 15%, and a stock that goes up 10%. 
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold'}


dates <- seq(1, 21)
stockA_values <- numeric(21)
stockA_values[1] <- 100
for (i in 2:21) {
  if (i %% 2 == 0) {  
    stockA_values[i] <- stockA_values[i-1] * 1.5
  } else { 
    stockA_values[i] <- stockA_values[i-1] * 0.85
  }
}
stockB_values <- 100
for (i in 2:21)
{
  stockB_values[i] <- stockB_values[i - 1] * 1.1 
}

stocks = data.frame(Date = dates, StockA = stockA_values, StockB = stockB_values)

ggplot(stocks) +
  geom_line(aes(x = Date, y = StockA, color = "StockA"), size = 1) +
  geom_line(aes(x = Date, y = StockB, color = "StockB"), size = 1) +
  labs(
    title = "StockA and StockB Over Time",
    x = "Time (Days)",
    y = "Price",
    color = "Stocks"  # Legends will show StockA and StockB
  ) +
  scale_color_manual(values = c("StockA" = "blue", "StockB" = "green")) +  # Custom colors for each stock
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )

stocks$StockA_3xlevered <- gen_daily_leveraged_returns(stocks$StockA, 3)
stocks$StockB_3xlevered <- gen_daily_leveraged_returns(stocks$StockB, 3)

ggplot(stocks) +
  geom_line(aes(x = Date, y = StockA_3xlevered, color = "StockA"), size = 1) +
  geom_line(aes(x = Date, y = StockB_3xlevered, color = "StockB"), size = 1) +
  labs(
    title = "StockA and StockB 3x Leveraged Over Time",
    x = "Time (Days)",
    y = "Price",
    color = "Stocks"  # Legends will show StockA and StockB
  ) +
  scale_color_manual(values = c("StockA" = "blue", "StockB" = "green")) +  # Custom colors for each stock
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )

```



Upon simulating the performance of these stocks, it becomes clear that while Stock A outperformed Stock B in absolute terms, the leveraged derivative of Stock B exhibited significantly superior performance. The simplest expiation why is that a 50% loss requires a 100% gain to make up, meanwhile a 50% gain only takes a 33.333% loss to destroy. This simple concept is extremely important when dealing with leveraged high risk investments.

Let's look at how this simple relationship plays out in our simulation. Stock A normally would have a cycle of a 50% gain followed by a 15% loss, which would result in an approximate 12.19%, which is better then Stock B, which only had a 10% compounding result. This means that we would expect that Stock A would have a 36.57% aproxmit compounding gain, while Stock B would only have 30%, but in reality a 150% gain and 45% loss only results in an approximate 17.26% compounding gain. This means that growth isn't the only thing that matters with leveraged products, but by how much it grows is also important. This is one of the main reasons why all leveraged products will say that they are only there for day traders as in the long run they will not have 3x returns. Now there are some advantages to this model, which is that circuit breakers will prevent the stock from ever hitting that 33.333% loss that is required to take our position to 0. Meanwhile, if we took leverage over a longer period lets say 6 months. If SPY fell by 33.333% over those 3 months our position would go to 0.

Now the worst part about this day to day resetting model isn't that losses stack up as if SPY falls by 20%, you already are expecting to lose a lot of money, but what happens during flat markets. 

To get an idea of what can happen lets simulate a stock that rises by 11% and falls by 10%. 



```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold'}

stockC <- numeric(20)

stockC <- 100

for (i in 2:20)
{
  if (i %% 2 == 0)
  {
    stockC[i] = stockC[i - 1] * 1.1
  }
  
  else
  {
    stockC[i] = stockC[i - 1] * 0.9
  }
}


stockC_leveraged = gen_daily_leveraged_returns(stockC, 3)

data <- data.frame(
  Day = 1:20,
  StockC = stockC,
  StockC_Leveraged = stockC_leveraged
)



ggplot(data, aes(x = Day)) +
  geom_line(aes(y = StockC, color = "StockC"), size = 1) +
  geom_line(aes(y = StockC_Leveraged, color = "Stock C Leveraged"), size = 1) +
  labs(
    title = "StockC vs StockC Leveraged",
    x = "Day",
    y = "Value",
    color = "Legend"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )




```

As you can see the underlying finishes at roughly 100, the leveraged version loses over 50% of its value during this time. In this regard, leveraged ETFs are kind of like an option where both volatility and price matters. 

# SPXTR vs SPXL

Let's compare the total return of the S&P 500 to the 3x leveraged S&P 500 daily. Since inception, which was post 2008 SPY has not had much volatility to the point where shorting the VIX was a popular strategy all the way up until Volmageddon, which was when several inverse VIX ETPs collapsed in 2018.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold'}



SPXL <- update_time_data(read.csv("SPXL.csv"), "SPXL")
SPXTR <- update_time_data(read.csv("^SP500TR.csv", header = TRUE), "SPXTR")




base_year <- 2008 + (11/12)
SPXL <- update_for_base_year(SPXL[SPXL$DecimalYear >= base_year, ], base_year)
SPXTR <- update_for_base_year(SPXTR[SPXTR$DecimalYear >= base_year, ], base_year)

SPXL$index <- SPXL$Open / SPXL$Open[1] * 100000
SPXTR$index <- SPXTR$Open / SPXTR$Open[1] * 100000

ggplot() +
  # Plot SPXL data (normalized)
  geom_line(data = SPXL, aes(x = DecimalYear, y = index, color = "SPXL"), size = 1) +
  
  geom_line(data = SPXTR, aes(x = DecimalYear, y = index, color = "SPXTR"), size = 1) +
  # Add labels and theme
  labs(
    title = "Log price of SPXL vs SPXTR",
    x = "Year",
    y = "Normalized Close Price (Base December 2008)",
    color = "ETF"
  ) +
  theme_minimal() +
  scale_color_manual(values = c( "SPXL" = "red", "SPXTR" = "green")) +
  theme(legend.position = "top") + scale_y_log10()


```


If you would have invested 100K into SPXL at inception, you would have quickly second guessed your option as it collapsed almost to 50K, but if you would have stuck around you would have 7,654,218, meanwhile if you would have just invested in the S&P 500 you would have had 1,011,590. 

As I stated before the simulation though, this happened during a time when shorting VIX was a popular stratgey. Now to see if its better then the S&P 500 it would be nice to stretch our simulation very far back as this would allow it to play over different periods. Luckily with Yahoo Finances Historical S&P 500, we can simulate from before the great depression. 



# The backtest of the centry


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold'}

SPY_data <- read.csv("^GSPC_daily.csv")
SPY_data <- update_time_data(SPY_data, "^GSPC")



SPY_data$med <- (SPY_data$High + SPY_data$Low) / 2 
SPY_data$index <- SPY_data$med / SPY_data$med[1] * 100
SPY_data$index_3x <- gen_daily_leveraged_returns(SPY_data$index, 3)





SPXTR_old <- read.csv("SPXTR_old.csv")

SPXTR_old$Price <- as.numeric(gsub(",", "", SPXTR_old$Amount))

SPXTR_old$DecimalYear <- SPXTR_old$Year + (SPXTR_old$Month - 1) / 12

SPXTR_old$index <-SPXTR_old$Price / SPXTR_old$Price[1] * 100






SPY_data$DecimalYear <- SPY_data$DecimalYear  + (as.numeric(substr(SPY_data$Date, 9, 10)) / 31)



beginning_value_index <- SPY_data$index[1]

ending_value_index <- SPY_data$index[nrow(SPY_data)]
distinct_count <- SPY_data %>% summarize(count = n_distinct(SPY_data$Year))

begginning_value_index_3x <- SPY_data$index_3x[1]
ending_value_index_3x <- SPY_data$index_3x[nrow(SPY_data)]

begginning_value_index_SPXTR <- SPXTR_old$index[1]
ending_value_index_SPXTR <- SPXTR_old$index[nrow(SPXTR_old)]


CAGR_index <- (ending_value_index - beginning_value_index) ^ (1 / distinct_count$count[1]) - 1
CAGR_index_3x <- (ending_value_index_3x - begginning_value_index_3x) ^ (1 /distinct_count$count[1]) - 1
CAGR_index_spxtr <- (ending_value_index_SPXTR - begginning_value_index_SPXTR) ^  (1 /distinct_count$count[1]) - 1

CAGR_index_label <- paste("S&P 500 (CAGR:", round(CAGR_index * 100, 2), "%)")
CAGR_index_3x_label <- paste("Leveraged S&P 500 (CAGR:", round(CAGR_index_3x * 100, 2), "%)")
CAGR_SPXTR_label <- paste("Total Return S&P (CAGR:", round(CAGR_index_spxtr * 100, 2), "%)")



ggplot() +
  geom_line(data = SPY_data, aes(x = DecimalYear, y = index_3x, color = "Leveraged S&P 500"), size = 1) + 
  geom_line(data = SPY_data, aes(x = DecimalYear, y = index, color = "S&P 500"), size = 1) +
  labs(
    title = "SPY Index vs SPXTR Index",
    x = "Year",
    y = "Index Value",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Leveraged S&P 500" = "blue", "S&P 500" = "red"),
                     labels = c(CAGR_index_3x_label, CAGR_index_label)) +
  theme_minimal()


ggplot() +
  geom_line(data = SPY_data, aes(x = DecimalYear, y = index_3x, color = "Leveraged S&P 500"), size = 1) + 
  geom_line(data = SPXTR_old, aes(x = DecimalYear, y = index, color = "S&P 500"), size = 1) +
  labs(
    title = "SPY Index vs SPXTR Index",
    x = "Year",
    y = "Index Value",
    color = "Legend"
  ) +
  scale_color_manual(
    values = c("Leveraged S&P 500" = "blue", "S&P 500" = "red"),
    labels = c(CAGR_index_3x_label, CAGR_SPXTR_label)
  ) +
  theme_minimal()


```

Looking at the first chart the leveraged S&P 500 looks extremely good as it has a 2% higher CAGR over a period of 100 years, but this model only accounts for price changes. When you invest in S&P 500 you get both the capital appreciation and the dividends, which then are reivested. Leveraged ETFs miss out on these dividends as they usually use futures for leverage. This means that, while you would have gotten a nice 13.28% CAGR, the S&P 500 would have had a 15.14% CAGR. 


# What if we pick something with low dividends

Our last simulation was not very good for leveraged ETFs, so instead lets look at QQQ. QQQ is the top 100 companies listed on the NASDAQ, which is tech heavy. Its dividend yield is about 0.5% where SPY is about 1.21%. 

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold'}

QQQ_daily <- read.csv("QQQ_daily.csv")

QQQ_3x <- gen_daily_leveraged_returns(QQQ_daily$Open, 3)


library(ggplot2)

# Assuming QQQ_daily and QQQ_3x are already defined
# Create a data frame for plotting
plot_data <- data.frame(
  Date = as.Date(QQQ_daily$Date),  # Convert Date column to Date type
  QQQ_Open = QQQ_daily$Open,
  QQQ_3x = QQQ_3x
)

# Create the plot
ggplot(plot_data, aes(x = Date)) +
  geom_line(aes(y = QQQ_Open, color = "QQQ Daily Open"), size = 1) +
  geom_line(aes(y = QQQ_3x, color = "QQQ 3x Leveraged"), size = 1) +
  labs(title = "QQQ Daily vs QQQ 3x Leveraged Returns",
       x = "Date",
       y = "Price",
       color = "Legend") +
  theme_minimal()

```


In this instance, the 3x leverage performance does well, but then collpases during the .com buble, and is never able to recover. This raises the question of whether lower leverage could have better retruns then higher leverage as you lose less capital during bad years. 

Let's compare a QQQ to leveraged ETFs that leverage between 1-5. 


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold'}

# Load necessary libraries
library(ggplot2)
library(dplyr)

# Function to calculate CAGR
calculate_cagr <- function(prices, n_years) {
  final_value <- tail(prices, 1)
  initial_value <- head(prices, 1)
  cagr <- ((final_value / initial_value)^(1 / n_years)) - 1
  return(cagr)
}

# Load data
QQQ_daily <- read.csv("QQQ_daily.csv")

# Calculate the number of years in the dataset
n_years <- as.numeric(difftime(max(as.Date(QQQ_daily$Date)), min(as.Date(QQQ_daily$Date)), units = "days")) / 365

# Generate leveraged returns for leverages 1 through 5 and calculate CAGR for each
cagrs <- c()
for (leverage in 1:5) {
  column_name <- paste0("QQQ_", leverage, "x")
  QQQ_daily[[column_name]] <- gen_daily_leveraged_returns(QQQ_daily$Open, leverage)
  cagrs <- c(cagrs, calculate_cagr(QQQ_daily[[column_name]], n_years))
}

# Prepare data for plotting
plot_data <- data.frame(
  Date = as.Date(QQQ_daily$Date),
  QQQ_Open = QQQ_daily$Open,
  QQQ_1x = QQQ_daily$QQQ_1x,
  QQQ_2x = QQQ_daily$QQQ_2x,
  QQQ_3x = QQQ_daily$QQQ_3x,
  QQQ_4x = QQQ_daily$QQQ_4x,
  QQQ_5x = QQQ_daily$QQQ_5x
)

# Create the plot
ggplot(plot_data, aes(x = Date)) +
  geom_line(aes(y = QQQ_Open, color = paste0("QQQ Daily Open (CAGR: ", round(cagrs[1] * 100, 2), "%)")), size = 1) +
  geom_line(aes(y = QQQ_1x, color = paste0("QQQ 1x Leveraged (CAGR: ", round(cagrs[1] * 100, 2), "%)")), size = 1) +
  geom_line(aes(y = QQQ_2x, color = paste0("QQQ 2x Leveraged (CAGR: ", round(cagrs[2] * 100, 2), "%)")), size = 1) +
  geom_line(aes(y = QQQ_3x, color = paste0("QQQ 3x Leveraged (CAGR: ", round(cagrs[3] * 100, 2), "%)")), size = 1) +
  geom_line(aes(y = QQQ_4x, color = paste0("QQQ 4x Leveraged (CAGR: ", round(cagrs[4] * 100, 2), "%)")), size = 1) +
  geom_line(aes(y = QQQ_5x, color = paste0("QQQ 5x Leveraged (CAGR: ", round(cagrs[5] * 100, 2), "%)")), size = 1) +
  labs(title = "QQQ Daily vs Leveraged Returns (1x to 5x)",
       x = "Date",
       y = "Price",
       color = "Legend") +
  theme_minimal()



```

The 2x leverage etf does the best followed by the 1x leverage, then the 3x leverage, and the rest finish near 0.  Now you may be wondering what happens if we do this with SPY. 


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold'}

# Load necessary libraries
library(ggplot2)
library(dplyr)

# Assuming SPY_data is already loaded
SPY_prices <- SPY_data$med

# Generate leveraged returns for SPY and store them
SPY_leveraged <- data.frame(Date = as.Date(SPY_data$Date))

for (leverage in 1:5) {
  column_name <- paste0("SPY_", leverage, "x")
  SPY_leveraged[[column_name]] <- gen_daily_leveraged_returns(SPY_prices, leverage)
}

# Create the plot without CAGR
ggplot(SPY_leveraged, aes(x = Date)) +
  geom_line(aes(y = SPY_1x, color = "SPY 1x Leveraged"), size = 1) +
  geom_line(aes(y = SPY_2x, color = "SPY 2x Leveraged"), size = 1) +
  geom_line(aes(y = SPY_3x, color = "SPY 3x Leveraged"), size = 1) +
  geom_line(aes(y = SPY_4x, color = "SPY 4x Leveraged"), size = 1) +
  geom_line(aes(y = SPY_5x, color = "SPY 5x Leveraged"), size = 1) +
  labs(title = "SPY Leveraged Returns (1x to 5x)",
       x = "Date",
       y = "Price",
       color = "Legend") +
  theme_minimal()


# Calculate daily returns for SPY_2x and SPY_3x
# Calculate daily returns for SPY_2x and SPY_3x
SPY_leveraged <- SPY_leveraged %>%
  mutate(
    SPY_2x_return = c(NA, diff(SPY_2x) / lag(SPY_2x)[-1]),
    SPY_3x_return = c(NA, diff(SPY_3x) / lag(SPY_3x)[-1])
  )

# Remove the NA values caused by the lag difference
returns_data <- SPY_leveraged %>%
  filter(!is.na(SPY_2x_return) & !is.na(SPY_3x_return))

# Perform the Wilcoxon test
wilcox_test_result <- wilcox.test(returns_data$SPY_2x_return, returns_data$SPY_3x_return)


wilcox_test_result <- wilcox.test(returns_data$SPY_2x_return, returns_data$SPY_3x_return)


cat("P-value:", wilcox_test_result$p.value, "\n")

```


As you can 


The 2x leveraged ETF slightly outperforms the 3x leveraged ETF. To verify whether the difference in returns is statistically significant, we performed a Wilcoxon test. The test yielded a very small p-value, indicating that the likelihood of the observed difference being due to random chance is extremely low. Therefore, it is highly probable that the returns are genuinely different.

# Conclusion
Over the long run, leveraged ETFs replicating the S&P 500 tend to under perform the regular S&P 500. This under performance is primarily due to missing out on the compounding effect of dividends. Consequently, leveraged ETFs should not replace regular investing but may serve as an alternative to long-term, out-of-the-money call options in a portfolio. However, as companies increasingly shift towards stock buybacks instead of paying dividends, leveraged ETFs could potentially become an effective way to maximize returns over an extended period.

